---
title: "Linear Mixed Models -- Curve Characteristics"
format: 
  html:
    self-contained: True 
editor: source
author: Brandon Kim, Andrew Kerr, Lily Cook, and Kyle Bistrain
code-fold: true
execute:
  message: false
  warning: false
---

## Mixed Effects Models and Visualizations for Elsa Bates

These code chunks work similarly to .jpynb code chunks.

Make sure to run this code chunk containing the required packages!

```{r}
library(tidyverse)
library(lmerTest)
library(emmeans)
library(multcomp)
```

```{r, warning = FALSE,message = FALSE, echo = TRUE}
# insert your local file path into the read_csv command
curve_characteristics <- read_csv("../data/curve_characteristics.csv")
head(curve_characteristics)
```

## Model with Response Proportion of time spent in a Harmful Posture

```{r}
# Fit a mixed-effects model:
#   Response: prop_harmful (proportion of time in harmful posture)
#   Predictors: movement, device, and their interaction
#   Random effects: intercept and slope for movement by participant
#   add a hashtag to comment out a line 
#   Comment out |> filter(movement != "lateral") if you want to include it in your analysis
prop_harmful_mod <- lmer(prop_harmful ~ movement*device + (1 + movement | participant),
                         data = curve_characteristics 
                                  |> filter(movement != "lateral")
                         )
```

```{r}
# evaluate model fit with residuals
plot(prop_harmful_mod)
```

```{r}
# Display summaries of fixed and random effects for each model
summary(prop_harmful_mod)
```

**Sample Interpretation of Fixed Effects**

-   Intercept (Estimate = 0.27575, p\<2e-16): This predicted proportion of time spent in harmful posture for the Axial movement measured with the IMU device is 0.27575.
-   movementflexion (Estimate = -0.01997, p = 0.405): Changing from axial to flexion movement with the IMU has a predicted decrease of 0.01997 in harmful posture proportion.
-   devicemocap (Estimate = 0.09160, p \< 1e-7): Under Axial movement, MoCap measures a predicted harmful-posture proportion that is 0.0916 higher than the IMU.
-   movementflexion:devicemocap (Estimate = -0.09939, p \< 1e-5): The interaction effect between flexion and mocap indicates for MoCap, the predicted proportion of time in a harmful posture is about 9.9939 percentage points lower during Flexion than it is during axial movement.

**Sample Interpretation of Random Effects**

-   Participant slope for Flexion (Variance = 0.00981, SD = 0.0990): Individuals differ more in how Flexion changes their posture proportion.
-   Correlation = –0.84 between intercept and slope demonstrates that participants with higher baseline proportion harmful posture tend to show smaller Flexion effects and vice versa.

```{r}
# analysis of variance
anova(prop_harmful_mod)
```

**Obtain estimates:**

```{r}
# Plot estimated marginal means (EMMs) for prop_harmful_mod
# CIs = TRUE adds confidence intervals
emmeans(prop_harmful_mod, specs = ~ device|movement)

emmeans(prop_harmful_mod, specs = ~ device|movement) |> 
  pairs(infer = c(T,T))
```

```{r}
# Plot estimated marginal means (EMMs) for prop_harmful_mod
# CIs = TRUE adds confidence intervals
emmip(prop_harmful_mod, device ~ movement, CIs = T) +
  theme_bw() +
  scale_color_manual(labels = c("IMU", "MoCap"), values = c("cornflowerblue", "darkgreen")) +
  scale_x_discrete(labels = c("axial" = "Axial", "flexion" = "Flexion")) +
  scale_y_continuous(limits = c(0,0.5)) +
  labs(y = "Proportion of Time in Harmful Posture",
       color = "Device") + 
  theme()
```


## Model with Response Estimated range of motion by Movement and device

```{r}
# Fit a mixed-effects model:
#   Response: Estimated range_of_motion (range of motion measured)
#   Predictors and random effects same as above
#   Comment out |> filter(movement != "lateral") if you want to include it in your analysis

range_of_motion_mod <- lmer(range_of_motion ~ movement*device + (1 + movement | participant),
                         data = curve_characteristics 
                                  |> filter(movement != "lateral")
                         )
```

```{r}
plot(range_of_motion_mod)
```

```{r}
# Display summaries of fixed and random effects
summary(range_of_motion_mod)
```

```{r}
anova(range_of_motion_mod)
```


```{r}
# Plot estimated marginal means (EMMs) for prop_harmful_mod
# CIs = TRUE adds confidence intervals
emmeans(range_of_motion_mod, specs = ~ device|movement)

emmeans(range_of_motion_mod, specs = ~ device|movement) |> 
  pairs(infer = c(T,T))
```

```{r}
# Plot estimated marginal means (EMMs) for range of motion
# CIs = TRUE adds confidence intervals
emmip(range_of_motion_mod, device ~ movement, CIs = T) +
  theme_bw() +
  scale_color_manual(labels = c("IMU", "MoCap"), values = c("cornflowerblue", "darkgreen")) +
  scale_x_discrete(labels = c("axial" = "Axial", "flexion" = "Flexion"))+
  labs(y = "Estimated Range of Motion",
       color = "Device") + 
  theme()
```

